\documentclass[11pt,openright]{book}

\usepackage[paperwidth=6.5in,paperheight=9.25in,top=0.78in,bottom=0.72in,inner=0.82in,outer=0.68in,headheight=23pt]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{amsmath,amssymb}
\usepackage{booktabs,tabularx,longtable,array}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,shapes.geometric,calc}
\usepackage{listings}
\usepackage{enumitem}
\usepackage{url}
\usepackage{makeidx}
\usepackage{fancyhdr}
\usepackage{hyperref}

\definecolor{ink}{HTML}{17212B}
\definecolor{muted}{HTML}{53616D}
\definecolor{accent}{HTML}{0B7285}
\definecolor{codebg}{HTML}{F3F6F8}
\definecolor{codekw}{HTML}{075985}
\definecolor{codecomment}{HTML}{4D7C0F}
\definecolor{rulegray}{HTML}{C9D2D9}

\hypersetup{
  colorlinks=true,
  linkcolor=accent,
  citecolor=accent,
  urlcolor=accent,
  pdftitle={Engineering Software},
  pdfauthor={Scott Brodie Forsyth}
}

\lstset{
  basicstyle=\ttfamily\small,
  backgroundcolor=\color{codebg},
  frame=single,
  rulecolor=\color{rulegray},
  framerule=0.4pt,
  framesep=5pt,
  xleftmargin=7pt,
  xrightmargin=7pt,
  breaklines=true,
  columns=fullflexible,
  keepspaces=true,
  showstringspaces=false,
  keywordstyle=\color{codekw}\bfseries,
  commentstyle=\color{codecomment}\itshape,
  numbers=left,
  numberstyle=\tiny\color{muted},
  numbersep=6pt
}

\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE]{\small\sffamily\color{muted}\leftmark}
\fancyhead[RO]{\small\sffamily\color{muted}\rightmark}
\fancyfoot[C]{\small\sffamily\thepage}
\setlength{\headwidth}{\textwidth}

\setlist{itemsep=2pt,topsep=4pt}
\setlength{\tabcolsep}{3pt}
\setcounter{tocdepth}{1}
\setcounter{secnumdepth}{2}
\makeindex

\makeatletter
\renewcommand{\cleardoublepage}{%
  \clearpage
  \if@twoside
    \ifodd\c@page
    \else
      \hbox{}\thispagestyle{empty}\newpage
      \if@twocolumn\hbox{}\newpage\fi
    \fi
  \fi
}
\makeatother

\newcommand{\term}[1]{\textbf{#1}\index{#1}}
\newcommand{\artifact}[1]{\texttt{#1}}
\newenvironment{keyidea}{\begin{quote}\small\itshape\color{accent}}{\end{quote}}
\newenvironment{tradeoffs}{\begin{quote}\small\color{muted}\textbf{Trade-offs.}\ }{\end{quote}}
\newenvironment{exercise}{\begin{quote}\small\color{ink}\textbf{Exercise.}\ }{\end{quote}}

\begin{document}

\frontmatter
\begin{titlepage}
  \pagecolor{ink}
  \color{white}
  \vspace*{1.15in}
  {\sffamily\fontsize{24}{28}\selectfont\bfseries ENGINEERING SOFTWARE\par}
  \vspace{0.22in}
  {\sffamily\Large From Reliable Code to Production Systems\par}
  \vspace{0.36in}
  \textcolor{white!55}{\rule{1.2in}{0.8pt}}\par
  \vfill
  {\sffamily\large Scott Brodie Forsyth\par}
  \vspace{0.25in}
  {\sffamily\small A compact, evidence-aware guide to designing, building,\par\noindent\hspace*{0pt}testing, securing, deploying, and operating software.\par}
  \thispagestyle{empty}
\end{titlepage}
\nopagecolor

\cleardoublepage
\thispagestyle{empty}
\vspace*{1.2in}
\begin{center}
  {\Large\sffamily\bfseries Engineering software is the practice of making\par}
  \vspace{0.15in}
  {\large\sffamily\color{muted}important behaviour dependable under changing conditions.\par}
\end{center}
\vfill
\noindent\small
Copyright \textcopyright\ 2026 Scott Brodie Forsyth.\par
\vspace{0.1in}
This edition is prepared as an editable educational manuscript. Standards,
dependencies, and platform interfaces change; verify operational details before
using them in a live system.\par
\vspace{0.1in}
Source code examples are provided for learning and are licensed for reuse with
the repository. Third-party standards and books remain under their original
licenses.

\cleardoublepage
\tableofcontents

\chapter*{How to use this book}
\addcontentsline{toc}{chapter}{How to use this book}
Software engineering is not a bag of rituals. It is a way of reducing the gap
between what a system is supposed to do and what it actually does when people,
networks, data, deadlines, attackers, and future maintainers interact with it.

Each chapter moves between four levels: a mental model, a concrete technique,
the conditions under which the technique helps, and the way it can fail. The
running project is a small order-intake service called \emph{Ledgerline}. It is
intentionally ordinary: it accepts orders, validates them, stores them, exposes
an API, emits useful signals, and must remain understandable six months later.
The repository contains smaller examples so that the ideas can be run without
proprietary infrastructure.

Words such as \emph{must}, \emph{should}, and \emph{may} are used carefully.
When they are capitalized inside a quoted standard, they retain the meaning of
that standard. Elsewhere, they indicate the strength of the author's advice,
not a universal law.

\mainmatter
\input{chapters/01-engineering}
\input{chapters/02-requirements}
\input{chapters/03-design}
\input{chapters/04-interfaces-data}
\input{chapters/05-concurrency-distributed}
\input{chapters/06-architecture}
\input{chapters/07-paradigms}
\input{chapters/08-collaboration}
\input{chapters/09-quality}
\input{chapters/10-reliability}
\input{chapters/11-security}
\input{chapters/12-delivery}
\input{chapters/13-operations}
\input{chapters/14-maintenance}
\input{chapters/15-human-systems}
\input{chapters/16-leading-projects}
\input{chapters/17-case-study}
\input{chapters/18-capstone}

\appendix
\input{chapters/appendix-repository}
\input{chapters/appendix-exercises}
\input{chapters/appendix-solutions}

\backmatter
\chapter*{Glossary}
\addcontentsline{toc}{chapter}{Glossary}
\footnotesize
\begin{description}[style=nextline,leftmargin=1.35in,labelwidth=1.2in]
\item[API] A defined interface through which software components exchange requests and results.
\item[Availability] The proportion of time a service is usable under a stated measurement definition.
\item[Capacity] The sustainable amount of work a system can handle while meeting its objectives.
\item[Coupling] The degree to which one component depends on another component's details or timing.
\item[Deployment] The act of making a version available in an execution environment.
\item[Idempotency] The property that repeating an operation has the same intended effect as performing it once.
\item[Invariant] A condition that remains true across a specified operation or state transition.
\item[Latency] The elapsed time between an operation's start and completion, measured at a defined boundary.
\item[Observability] The ability to infer internal system state from externally emitted signals.
\item[Recovery point objective] The maximum acceptable amount of data loss measured in time.
\item[Recovery time objective] The maximum acceptable time to restore a service after a failure.
\item[Reliability] The probability that a system performs its required function for a stated period and context.
\item[Service-level objective] A target for a measurable service indicator over a stated window.
\item[Technical debt] A present design or implementation choice whose future cost is knowingly incurred to gain a short-term benefit.
\item[Threat model] A structured description of assets, actors, trust boundaries, threats, and mitigations.
\end{description}

\bibliographystyle{plain}
\bibliography{references}

\printindex
\end{document}
