\chapter{Selected solutions and reasoning}

\section{Foundations}
\begin{enumerate}
  \item A bicycle rental boundary can include reservation, payment authorization,
        and lock commands. The rider, payment provider, and physical lock are
        outside it. Invariants include one active rental per bicycle, a return
        cannot precede a start, and a charged amount matches the tariff snapshot.
  \item “Fast” becomes a journey, a percentile, a window, and a threshold. For
        example: p95 from submit to visible confirmation is below 400 ms for a
        defined payload at a stated load. Add invalid-input and dependency-failure
        acceptance cases.
  \item A notification diagram should show the caller, notification service,
        provider, delivery callback, and operator. The provider is a trust and
        reliability boundary; callbacks need authentication and replay handling.
\end{enumerate}

\section{Design and data}
\begin{enumerate}
  \setcounter{enumi}{4}
  \item Useful modules for document import are format parsing, validation,
        canonical domain model, persistence, and reporting. The parser should not
        decide authorization or write directly to the database.
  \item A cursor should encode or reference a stable ordering key and direction.
        It must not assume that page number remains stable while rows are inserted.
  \item A subscription state machine might allow trial -> active -> paused ->
        cancelled, but disallow cancelled -> active without a new subscription or
        explicit reactivation policy. The rule belongs in one transition function.
  \item A foreign key with `ON DELETE RESTRICT` prevents deletion while children
        exist; `ON DELETE CASCADE` is appropriate only when child loss is intended.
  \item Add the new field, write both fields, read the new field with fallback,
        backfill, enforce the new invariant, then stop writing and remove the old
        field after the compatibility window.
\end{enumerate}

\section{Concurrency and architecture}
\begin{enumerate}
  \setcounter{enumi}{9}
  \item Two reservations can both read stock 1. An atomic update such as
        `UPDATE stock SET available = available - 1 WHERE sku = ? AND available > 0`
        followed by checking affected rows prevents the lost update.
  \item A consumer acknowledges only after its effect is durable. If it crashes
        before acknowledgment, the message may repeat; use a unique event ID or
        idempotent upsert and accept that exactly-once effects are not implied.
  \item With three engineers and modest traffic, a modular monolith minimizes
        operational and coordination cost. A service split becomes justified by
        distinct ownership, legal isolation, or a measured scaling boundary.
  \item An ADR should name workload, transactions, operational skill, migration
        cost, and alternatives. “SQL is standard” is not enough evidence.
  \item A dependency rule, migration coexistence test, and tenant-isolation test
        are three possible fitness functions.
\end{enumerate}

\section{Quality and operation}
\begin{enumerate}
  \setcounter{enumi}{14}
  \item Cover authenticated allowed access, authenticated denied access,
        unauthenticated access, malformed identifiers, oversized input, and
        dependency failure. State which combinations are intentionally omitted.
  \item For every valid item list, total equals the sum of positive integer
        quantity-times-price products and remains within the chosen numeric bound.
  \item An SLO must identify eligible requests, success definition, latency
        boundary, window, and excluded maintenance if any. Its budget is the
        complement of the target over the eligible measurement population.
  \item First stop or rate-limit intake if backlog threatens data freshness, then
        inspect consumer errors, dependency latency, and poison messages before
        increasing concurrency.
  \item A postmortem should preserve evidence, identify the deletion path and
        missing guard, restore from a verified backup, and add a migration check
        plus permission or review control. Blaming the operator is not a control.
\end{enumerate}

\section{Security and human impact}
\begin{enumerate}
  \setcounter{enumi}{19}
  \item Use short-lived, single-purpose reset tokens stored as hashes, bind them
        to the account and intent, invalidate on use, rate-limit attempts, avoid
        account enumeration, and log safe audit events.
  \item Remove secrets and unnecessary personal data from telemetry; define access,
        retention, deletion, and break-glass review for remaining records.
  \item Check semantic controls, focus order, visible focus, labels, error
        announcements, and completion without a pointer device. Confirm with
        assistive technology rather than relying on a visual inspection.
  \item Store canonical instants, currency, locale, and user preferences
        separately. Localize format at the presentation boundary.
  \item Verify every claimed API and dependency, run tests and security checks,
        inspect data handling, and require a human owner for approval and release.
\end{enumerate}

\section{Capstone reasoning}
\begin{enumerate}
  \setcounter{enumi}{24}
  \item An outbox row and the order state belong in one local transaction. The
        consumer uses a unique event ID and records completion. Re-delivery is
        expected; the external effect must be idempotent or reconciled.
  \item A migration should be rerunnable or safely resumable, report rejected
        rows, compare source and target totals, and stop before destructive work
        if reconciliation fails.
  \item A canary needs a cohort, success indicators, a time window, an owner,
        and an automatic or human-controlled rollback action. Do not expose a
        wider audience before telemetry is healthy.
  \item Measure a realistic workload and identify the saturated resource. A
        higher average throughput number without tail and error data is incomplete.
  \item The best architecture review names deferred features and the signals that
        would justify them. Restraint is part of design.
\end{enumerate}

